Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-11-12 manuscript/manuscript.tex
Fix wrong substitution of "\centreing"
Added iftwocol to grid-search figure in manuscript.tex
Fixed typo in author name in manuscript.tex
Fix editor's remark 5

\title[Variable Density Tesseroids]{
    Gravitational field calculation in spherical coordinates using variable
    densities in depth
}
\author[S.R. Soler, A. Pesce, L. Uieda and M.E. Gimenez]{
    Santiado R. Soler$^{1,2}$, Agustina Pesce$^{1,2}$, Leonardo Uieda$^3$ and
    Mario E. Gimenez$^{1,2}$ \\
    $^1$CONICET, Argentina. e-mail: santiago.r.soler@gmail.com\\
    $^2$Instituto Geofísico Sismológico Volponi, Universidad Nacional de
    San Juan, Argentina\\
    $^3$Department of Earth Sciences, SOEST, University of Hawai‘i at
    M\={a}noa, Honolulu, Hawaii, USA
}

@@

\begin{figure}
\centreing
\includegraphics[width=0.6\linewidth]{figures/tesseroid-uieda.pdf}
\caption{
    A tesseroid (spherical prism) in a geocentric spherical coordinate system, with a
    computation point $P$ and its local north oriented Cartesian coordinate system.
    After \citet{Uieda2015}.
}
\label{fig:tesseroid}
\end{figure}

The main challenge of the GLQ integration is loss of accuracy when the computation point
gets closer to the tesseroid \citep{Ku1977}.
\citet{Uieda2016} built on the adaptive discretization algorithm of \citet{Li2011}
to automatically obtain integration results with 0.1\% accuracy.
The algorithm consist in recursively splitting the tesseroid into smaller ones when a
threshold is exceeded,
namely when the normalized distance to the computation point is greater than a
``distance-size ratio'' ($D$).
\citet{Uieda2016} have also obtained standard values of $D$
for the gravitational potential, acceleration, and gradient tensor components
by comparing the numerical model with the fields generated by a spherical shell.

@@

\begin{figure*}
\centreing
\includegraphics[width=\linewidth]
    {figures/density-based-discretization-algorithm.pdf}
\caption{
    Example application of the density-based discretization algorithm to a non-linear
    density function.
    (a) the normalised density function $\rho_n(r')$ (blue), current boundaries of the
    tesseroid (orange dots), and the linear density function $\rho_l(r')$ (orange line).
    The dashed red line represents the maximum density difference $\Delta \rho (r')$ at
    which the tesseroid would be divided (assuming that the inequality
    \ref{eq:delta-density} is not satisfied).
    (b) second iteration of the algorithm with a new linear density function and maximum
    density difference. The tesseroid would be divided at the depth indicated by the
    dashed red line.
    (c) third iteration of the algorithm.
    (d) final output of the density-based discretization, assuming that all four new
    tesseroids satisfy inequality \ref{eq:delta-density}.
}
\label{fig:density-discretization-algorithm}
\end{figure*}

@@

\begin{figure*}
\centreing
\includegraphics[width=\linewidth]{figures/linear-D.pdf}
\caption{
    Differences between the gravitational fields generated by the tesseroid model
    and the analytical solution for a thin (left) and a thick (right) spherical shell.
    Both models have a linear density function of radius (Eq. \ref{eq:density-linear}).
    The computations were performed on the four combinations described in
    Table \ref{tab:grids}.
    Due to the linearity of the density function, the density-based discretization
    algorithm is not applied.
    Differences are reported as a percentage of the shell values.
    The horizontal dashed black line represents a target difference of 0.1\%.
}
\label{fig:D-linear}
\end{figure*}

@@

\begin{figure}
\centreing
\includegraphics[width=0.5\linewidth]
    {figures/grid-search.pdf}
\caption{
    Numerical error exploration in the $D$-$\delta$ space.
    The percentage difference values were obtained from the comparison
    between the analytical solution and the numerical approximation of
    the gravitational fields ($V$, $g_z$ and $g_{zz}$) generated by a
    spherical shell with an exponential density function (Eq. \ref{eq:density-exp}).
    These comparisons were carried out on the ``Big Grid'' (Table
    \ref{tab:grids}), with a spherical shell of
    35km of thickness, and density function with $b$ of 1km.
    The points inside the dashed line are the ones that present an
    error lower than 0.1\%.
    }
\label{fig:grid-search}
\end{figure}

@@

\begin{figure*}
\centreing
\includegraphics[width=\linewidth]{figures/exponential-delta.pdf}
\caption{
    Numerical error for different exponential density functions and values of the delta
    ratio.
    The top panels show the exponential density functions for different values of $b$
    (Eq. \ref{eq:density-exp}).
    The error computations were performed on the ``Big Grid'' configuration described in
    Table \ref{tab:grids}, with fixed values of the distance-size ratio
    $D$ obtained for the linear density case ($D=1, 2, 8$ for the
    potential, gradient components, and tensor components, respectively).
    The difference is reported as a percentage of the shell values.
    }
\label{fig:delta-exponential}
\end{figure*}

@@

\begin{figure*}
\centreing
\includegraphics[width=\linewidth]{figures/neuquen-basin.pdf}
\caption{
    Gravitational effects of the Nequ\'en sedimentary basin modelled
    using tesseroids with an exponential density function of depth.
    (a) Topography of the Neuqu\'en Basin (in km) and its location in South America,
    (b) thickness of the sedimentary basin \citep[in meters;][]{Heine2007},
    (c) the exponential density function used to model the basin
        (depth in km and density in kg/m$^3$),
    (d)-(l) computed gravitational fields: potential $V$ (in J/kg), gradient
    components $g_x$, $g_y$ and $g_z$ (in mGal), and Marussi tensor components
    (in Eötvös), calculated at 50km of height over the ellipsoid.
}
\label{fig:neuquen-basin}
\end{figure*}

@@

\begin{figure}
\centreing
\includegraphics[width=0.5\linewidth]{figures/spherical-shell.pdf}
\caption{
    Spherical shell with inner and outer radii $R_1$ and $R_2$, respectively.
    The computation point $Q$ is located on the $z$ axis at a distance $r$ from
    the origin of the coordinates system.
    For our purposes we will assume that $Q$ is outside of the shell,
    i.e. $r > R_2$.
}
\label{fig:spherical-shell}
\end{figure}

>>>>

\title[Variable Density Tesseroids]{
    Gravitational field calculation in spherical coordinates using variable
    densities in depth
}
\author[S.R. Soler, A. Pesce, L. Uieda and M.E. Gimenez]{
    Santiago R. Soler$^{1,2}$, Agustina Pesce$^{1,2}$, Leonardo Uieda$^3$ and
    Mario E. Gimenez$^{1,2}$ \\
    $^1$CONICET, Argentina. e-mail: santiago.r.soler@gmail.com\\
    $^2$Instituto Geofísico Sismológico Volponi, Universidad Nacional de
    San Juan, Argentina\\
    $^3$Department of Earth Sciences, SOEST, University of Hawai‘i at
    M\={a}noa, Honolulu, Hawaii, USA
}

@@

\begin{figure}
\centering
\includegraphics[width=0.6\linewidth]{figures/tesseroid-uieda.pdf}
\caption{
    A tesseroid (spherical prism) in a geocentric spherical coordinate system, with a
    computation point $P$ and its local north oriented Cartesian coordinate system.
    After \citet{Uieda2015}.
}
\label{fig:tesseroid}
\end{figure}

The main challenge of the GLQ integration is loss of accuracy when the computation point
gets closer to the tesseroid \citep{Ku1977}.
\citet{Uieda2016} built on the adaptive discretization algorithm of \citet{Li2011}
to automatically obtain integration results with 0.1\% accuracy.
The algorithm consist in recursively splitting the tesseroid into smaller ones when a
threshold is exceeded,
namely when the normalized distance to the computation point is greater than a
``distance-size ratio'' ($D$).
\citet{Uieda2016} have also obtained standard values of $D$
for the gravitational potential, acceleration, and gradient tensor components
by comparing the numerical model with the analytical solution of the fields generated by
a spherical shell.

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]
    {figures/density-based-discretization-algorithm.pdf}
\caption{
    Example application of the density-based discretization algorithm to a non-linear
    density function.
    (a) the normalised density function $\rho_n(r')$ (blue), current boundaries of the
    tesseroid (orange dots), and the linear density function $\rho_l(r')$ (orange line).
    The dashed red line represents the maximum density difference $\Delta \rho (r')$ at
    which the tesseroid would be divided (assuming that the inequality
    \ref{eq:delta-density} is not satisfied).
    (b) second iteration of the algorithm with a new linear density function and maximum
    density difference. The tesseroid would be divided at the depth indicated by the
    dashed red line.
    (c) third iteration of the algorithm.
    (d) final output of the density-based discretization, assuming that all four new
    tesseroids satisfy inequality \ref{eq:delta-density}.
}
\label{fig:density-discretization-algorithm}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/linear-D.pdf}
\caption{
    Differences between the gravitational fields generated by the tesseroid model
    and the analytical solution for a thin (left) and a thick (right) spherical shell.
    Both models have a linear density function of radius (Eq. \ref{eq:density-linear}).
    The computations were performed on the four combinations described in
    Table \ref{tab:grids}.
    Due to the linearity of the density function, the density-based discretization
    algorithm is not applied.
    Differences are reported as a percentage of the shell values.
    The horizontal dashed black line represents a target difference of 0.1\%.
}
\label{fig:D-linear}
\end{figure*}

@@

\begin{figure}
\centering
\iftwocol{
\includegraphics[width=0.9\linewidth]
    {figures/grid-search.pdf}
}{
\includegraphics[width=0.5\linewidth]
    {figures/grid-search.pdf}
}
\caption{
    Numerical error exploration in the $D$-$\delta$ space.
    The percentage difference values were obtained from the comparison
    between the analytical solution and the numerical approximation of
    the gravitational fields ($V$, $g_z$ and $g_{zz}$) generated by a
    spherical shell with an exponential density function (Eq. \ref{eq:density-exp}).
    These comparisons were carried out on the ``Big Grid'' (Table
    \ref{tab:grids}), with a spherical shell of
    35km of thickness, and density function with $b$ of 1km.
    The points inside the dashed line are the ones that present an
    error lower than 0.1\%.
    }
\label{fig:grid-search}
\end{figure}

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/exponential-delta.pdf}
\caption{
    Numerical error for different exponential density functions and values of the delta
    ratio.
    The top panels show the exponential density functions for different values of $b$
    (Eq. \ref{eq:density-exp}).
    The error computations were performed on the ``Big Grid'' configuration described in
    Table \ref{tab:grids}, with fixed values of the distance-size ratio
    $D$ obtained for the linear density case ($D=1, 2, 8$ for the
    potential, gradient components, and tensor components, respectively).
    The difference is reported as a percentage of the shell values.
    }
\label{fig:delta-exponential}
\end{figure*}

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/neuquen-basin.pdf}
\caption{
    Gravitational effects of the Nequ\'en sedimentary basin modelled
    using tesseroids with an exponential density function of depth.
    (a) Topography of the Neuqu\'en Basin (in km) and its location in South America,
    (b) thickness of the sedimentary basin \citep[in meters;][]{Heine2007},
    (c) the exponential density function used to model the basin
        (depth in km and density in kg/m$^3$),
    (d)-(l) computed gravitational fields: potential $V$ (in J/kg), gradient
    components $g_x$, $g_y$ and $g_z$ (in mGal), and Marussi tensor components
    (in Eötvös), calculated at 50km of height over the ellipsoid.
}
\label{fig:neuquen-basin}
\end{figure*}

@@

\begin{figure}
\centering
\includegraphics[width=0.5\linewidth]{figures/spherical-shell.pdf}
\caption{
    Spherical shell with inner and outer radii $R_1$ and $R_2$, respectively.
    The computation point $Q$ is located on the $z$ axis at a distance $r$ from
    the origin of the coordinates system.
    For our purposes we will assume that $Q$ is outside of the shell,
    i.e. $r > R_2$.
}
\label{fig:spherical-shell}
\end{figure}
